Finally Tagless Staged Interpreters for Simpler Typed Languages

### 2007-04-06 tagless/tagless-final.tex
remove all comments.  make all coalgebra references use the same spelling.  check uses of initial/final (all ok)

In sum, by Church\hyp encoding terms using ordinary functions,
we achieve a tagless evaluator
for a typed object language in a metalanguage with a simple type
system \citep{hindley-principal,milner-theory}.  In this \emph{final}
rather than \emph{initial} approach, both kinds of run-time errors
in \S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. 
\oleg{Jacques, do we use final and initial here correctly?}
Because the new interpreter uses no
universal type or
pattern matching, it never results in a run-time error (and is in fact
total).  Because this safety is obvious not just to us but also to the
metalanguage implementation, we avoid the serious performance penalty
\citep{WalidICFP02} of error checking.  \Citet{Gluck-jones-optimality}
explains deeper technical reasons that inevitably lead to these performance
penalties.

@@

Our partial evaluator is optimal with respect to the self\hyp
interpreter~$\si$.
\begin{prop}
    Let $\pe$ be the partial evaluator~|P| in~\S\ref{S:PE-solution}.
    Then the object terms $\pe(\si[\encode{e}])$ and~$\pe(e)$ are
    either both undefined or both defined and
    equal up to $\alpha$\hyp conversion.
    \ccshan{Check this}
\end{prop}

@@

or, if we denote the sequence of initial arguments as S r, terms have
the type |S r -> r tau|
The interpreter has the type
|(forall r. S r -> r tau) -> r' tau|

@@

We have been unable to find a direct reference that establishes that
the \emph{typed} $\lambda$-calculus possesses a final co-algebra structure.
For the untyped $\lambda$-calculus, 
\citet{HonsellLenisa,honsell99coinductive} establish several interesting
results along this line.  Of particular interest is their use of
contexts with a hole \citep[p.13]{honsell99coinductive} for the definition
of \emph{observational equivalence}, which directly relates to our
use of terms with a hole as a context for our self-interpreter 
(\S\ref{selfinterp}).  The reader is directed to bibliography of
\citet{honsell99coinductive} for references to the foundational work
in this important area.  Particularly intriguing is the link to the
coinductive aspects of B\"{o}hm trees, as pointed out by
\citet{berarducci-models} and \citet[Example 4.3.4]{jacobs-coalgebra}.

@@

\acks%\subsection*{Acknowledgments}
We thank Martin Sulzmann and Walid Taha 
for helpful discussions.  Sam Staton, Pieter Hofstra, and Bart Jacobs
kindly provided some useful references on co-algebraic structures in the
$\lambda$\hyp calculus

>>>>

In sum, by Church\hyp encoding terms using ordinary functions,
we achieve a tagless evaluator
for a typed object language in a metalanguage with a simple type
system \citep{hindley-principal,milner-theory}.  In this \emph{final}
rather than \emph{initial} approach, both kinds of run-time errors
in \S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. 
Because the new interpreter uses no
universal type or
pattern matching, it never results in a run-time error (and is in fact
total).  Because this safety is obvious not just to us but also to the
metalanguage implementation, we avoid the serious performance penalty
\citep{WalidICFP02} of error checking.  \Citet{Gluck-jones-optimality}
explains deeper technical reasons that inevitably lead to these performance
penalties.

@@

Our partial evaluator is optimal with respect to the self\hyp
interpreter~$\si$.
\begin{prop}
    Let $\pe$ be the partial evaluator~|P| in~\S\ref{S:PE-solution}.
    Then the object terms $\pe(\si[\encode{e}])$ and~$\pe(e)$ are
    either both undefined or both defined and
    equal up to $\alpha$\hyp conversion.
%    \ccshan{Check this}
%    \jacques{seems correct.  If it said that the results are operationally
%    equivalent to e itself, then that would be less clear.}
\end{prop}

@@

or, if we denote the sequence of initial arguments as |S r|, terms have
the type |S r -> r tau|
The interpreter has the type
|(forall r. S r -> r tau) -> r' tau|

@@

We have been unable to find a direct reference that establishes that
the \emph{typed} $\lambda$-calculus possesses a final coalgebra structure.
For the untyped $\lambda$-calculus, 
\citet{HonsellLenisa,honsell99coinductive} establish several interesting
results along this line.  Of particular interest is their use of
contexts with a hole \citep[p.13]{honsell99coinductive} for the definition
of \emph{observational equivalence}, which directly relates to our
use of terms with a hole as a context for our self-interpreter 
(\S\ref{selfinterp}).  The reader is directed to bibliography of
\citet{honsell99coinductive} for references to the foundational work
in this important area.  Particularly intriguing is the link to the
coinductive aspects of B\"{o}hm trees, as pointed out by
\citet{berarducci-models} and \citet[Example 4.3.4]{jacobs-coalgebra}.

@@

\acks%\subsection*{Acknowledgments}
We thank Martin Sulzmann and Walid Taha 
for helpful discussions.  Sam Staton, Pieter Hofstra, and Bart Jacobs
kindly provided some useful references on coalgebraic structures in the
$\lambda$\hyp calculus
